---
title: "Plants of TO"
author: "Garland Xie"
date: "09/12/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r library}
library(rdryad)
library(brranching)
library(readxl)
library(dplyr)
library(janitor)
library(ggplot2)
library(forcats)
library(gghighlight)
```

# Import 

```{r}
# download from DRYAD
plants_link <- dryad_download(dois = "10.5061/dryad.1ns1rn8sg")

# read csv file 
plants <- read.csv(plants_link[[1]])
```

```{r}
# download from DRYAD
traits_link <- dryad_download(
  dois = "10.5061/dryad.6rn879c"
  )

# read csv file 
traits_leaf <- read_excel(
  traits_link[[1]][1], 
  sheet = "Traits 1"
      )

traits_stature <- read_excel(
  traits_link[[1]][1], 
  sheet = "Traits 2"
      )

traits_CN <- read_excel(
  traits_link[[1]][1],
  sheet = "Traits 3"
)
```

# Data cleaning

Unit measurements for each plant functional trait:

* Leaf dry matter content (unitless)
* Specific leaf area (mm-squared/mg)
* Plant height (m)
* Stem width (mm)
* Number of leaves (unitless)
* Leaf N content (%)
* Leaf C content (%)
* Leaf C:N (unitless)


```{r clean leaf traits for each site}
leaf_tidy <- traits_leaf %>%
  clean_names() %>%
  select(
    species_code, 
    site_code,
    site_type,
    replicate = replicate_number,
    ldmc, 
    sla = sla_mm_2_mg
  ) %>%
  filter(site_type == "meadow") %>%
  group_by(species_code, site_code) %>%
  summarize(mean_ldmc = mean(ldmc, na.rm = TRUE),
            sd_ldm = sd(ldmc, na.rm = TRUE),
            mean_sla = mean(sla, na.rm = TRUE),
            sd_sla = sd(sla, na.rm = TRUE),
            n = n())
  
leaf_tidy
```

```{r clean leaf traits across all sites}
leaf_tidy2 <- traits_leaf %>%
  clean_names() %>%
  select(
    species_code, 
    site_code,
    site_type,
    replicate = replicate_number,
    ldmc, 
    sla = sla_mm_2_mg
  ) %>%
  filter(site_type == "meadow") %>%
  group_by(species_code) %>%
  summarize(mean_ldmc = mean(ldmc, na.rm = TRUE),
            sd_ldm = sd(ldmc, na.rm = TRUE),
            mean_sla = mean(sla, na.rm = TRUE),
            sd_sla = sd(sla, na.rm = TRUE),
            n = n())
  
leaf_tidy2
```

```{r clean stature for each site}
stature_tidy <- traits_stature %>%
  clean_names() %>%
  filter(site_type == "meadow") %>%
  select(
    species_code, 
    site_code,
    site_type,
    height = height_m,
    stem_width = stem_width_mm, 
    num_leaves = number_of_leaves) %>%
  group_by(species_code, site_code) %>%
  summarize(
    mean_height = mean(height, na.rm = TRUE),
    sd_height = sd(height, na.rm = TRUE),
    mean_stem = mean(stem_width, na.rm = TRUE),
    sd_stem = sd(stem_width, na.rm = TRUE),
    mean_num_leaves = mean(num_leaves, na.rm = TRUE),
    sd_num_leaves = sd(num_leaves, na.rm = TRUE),
    rep = n()
  ) 

stature_tidy
```

```{r clean stature traits across all sites}
stature_tidy2 <- traits_stature %>%
  clean_names() %>%
  filter(site_type == "meadow") %>%
  select(
    species_code, 
    height = height_m,
    stem_width = stem_width_mm, 
    num_leaves = number_of_leaves) %>%
  group_by(species_code) %>%
  summarize(
    mean_height = mean(height, na.rm = TRUE),
    sd_height = sd(height, na.rm = TRUE),
    mean_stem = mean(stem_width, na.rm = TRUE),
    sd_stem = sd(stem_width, na.rm = TRUE),
    mean_num_leaves = mean(num_leaves, na.rm = TRUE),
    sd_num_leaves = sd(num_leaves, na.rm = TRUE),
    rep = n()
  ) 

stature_tidy2
```


```{r clean CN for each site}
CN_tidy <- traits_CN %>%
  clean_names() %>%
  filter(site_type == "meadow") %>%
  select(
    species_code = species, 
    site_code, 
    site_type, 
    leaf_carbon_percent, 
    leaf_nitrogen_percent) %>%
  group_by(species_code, site_code) %>%
  summarize(
    mean_leaf_C = mean(leaf_carbon_percent, na.rm = TRUE),
    sd_leaf_C = sd(leaf_carbon_percent, na.rm = TRUE),
    mean_leaf_N = mean(leaf_nitrogen_percent, na.rm = TRUE),
    sd_leaf_N = sd(leaf_nitrogen_percent, na.rm = TRUE), 
    n = n()) 

CN_tidy
```

```{r clean CN traits across all sites}
CN_tidy2 <- traits_CN %>%
  clean_names() %>%
  filter(site_type == "meadow") %>%
  select(
    species_code = species, 
    site_code, 
    leaf_carbon_percent, 
    leaf_nitrogen_percent) %>%
  group_by(species_code) %>%
  summarize(
    mean_leaf_C = mean(leaf_carbon_percent, na.rm = TRUE),
    sd_leaf_C = sd(leaf_carbon_percent, na.rm = TRUE),
    mean_leaf_N = mean(leaf_nitrogen_percent, na.rm = TRUE),
    sd_leaf_N = sd(leaf_nitrogen_percent, na.rm = TRUE), 
    n = n()) 

CN_tidy2
```

# Exploratory analysis: summary statistics

## Plant height 

```{r plot mean height at each meadow site, fig.height=6, fig.width=5, message=FALSE, warning=FALSE}

stature_tidy %>%
  mutate(
    mean_height = mean_height * 100,
    sd_height = sd_height * 100) %>%
ggplot(aes(y = species_code, x = mean_height)) +
  geom_point() + 
  gghighlight(species_code == "VIRO") + 
  facet_wrap(~site_code) + 
  labs(x = "Mean height (cm)", y = NULL) +
  theme_bw() +
  theme(axis.text.y = element_blank())
```


```{r plot mean height across all sites, fig.height=4, fig.width=4}

stature_tidy2 %>%
  mutate(
    mean_height = mean_height * 100,
    sd_height = sd_height * 100,
    species_code = factor(species_code),
    species_code = fct_reorder(species_code, mean_height)) %>%
ggplot(aes(y = species_code, x = mean_height)) +
  geom_point() + 
  gghighlight(species_code == "VIRO") + 
  labs(x = "Mean height (cm)", y = NULL) +
  theme_bw() 
```

## SLA


```{r plot mean LDMC at each meadow site, fig.height=4, fig.width=4, message=FALSE}
leaf_tidy %>%
ggplot(aes(y = species_code, x = mean_sla)) +  
  geom_point() +
  facet_wrap(~site_code) + 
  gghighlight(species_code == "VIRO") + 
  labs(x = "Mean specific leaf area (mm-squared per mg)", y = NULL) +
  theme_bw() +
  theme(axis.text.y = element_blank())
```
```{r plot mean height across all sites, fig.height=4, fig.width=4}

leaf_tidy2 %>%
  mutate(species_code = factor(species_code),
         species_code = fct_reorder(species_code, mean_sla)) %>%
ggplot(aes(y = species_code, x = mean_sla)) +
  geom_point() + 
  gghighlight(species_code == "VIRO") + 
  labs(x = "Mean specific leaf area (mm-squared per mg)", y = NULL) +
  theme_bw() 
```

## LDMC

```{r plot mean LDMC at each meadow site, fig.height=5, fig.width=4, message=FALSE}
leaf_tidy %>%
ggplot(aes(y = species_code, x = mean_ldmc)) +  
  geom_point() +
  facet_wrap(~site_code) + 
  gghighlight(species_code == "VIRO") + 
  labs(x = "Mean leaf dry matter content", y = NULL) +
  theme_bw() +
  theme(axis.text.y = element_blank())
```

```{r plot mean LDMC across all meadow sites, fig.height=4, fig.width=4}
leaf_tidy2 %>%
  mutate(species_code = factor(species_code),
         species_code = fct_reorder(species_code, mean_ldmc, .desc = TRUE)) %>%
  ggplot(aes(y = species_code, x = mean_ldmc)) +  
  geom_point() +
  gghighlight(species_code == "VIRO") + 
  labs(x = "Mean leaf dry matter content", y = NULL) +
  theme_bw() 
```

## Leaf C content 

```{r plot mean leaf C at each site, fig.height=5, fig.width=5, message=FALSE}
CN_tidy %>% 
ggplot(aes(y = species_code, x = mean_leaf_C)) +
  geom_point() +
  gghighlight(species_code == "VIRO") + 
  facet_wrap(~site_code) +
  labs(x = "Mean leaf C content (%)", y = NULL) + 
  theme_bw() +
  theme(axis.text.y = element_blank())
```

```{r plot mean leaf C across all sites, fig.height=5, fig.width=5, message=FALSE}
CN_tidy2 %>% 
  mutate(species_code = factor(species_code),
         species_code = fct_reorder(species_code, mean_leaf_C)) %>%
ggplot(aes(y = species_code, x = mean_leaf_C)) +
  geom_point() +
  gghighlight(species_code == "VIRO") + 
  labs(x = "Mean leaf C content (%)", y = NULL) + 
  theme_bw() 
```

## Leaf N content

```{r plot mean leaf N at each site, fig.height=4, fig.width=4, message=FALSE}
CN_tidy %>% 
ggplot(aes(y = species_code, x = mean_leaf_N)) +
  geom_point() +
  gghighlight(species_code == "VIRO") + 
  facet_wrap(~site_code) +
  labs(x = "Mean leaf N content (%)", y = NULL) + 
  theme_bw() +
  theme(axis.text.y = element_blank())
```

```{r plot mean leaf N across all sites, fig.height=4, fig.width=4, message=FALSE}
CN_tidy2 %>% 
  mutate(species_code = factor(species_code),
         species_code = fct_reorder(species_code, mean_leaf_N)
         ) %>%
  ggplot(aes(y = species_code, x = mean_leaf_N)) +
  geom_point() +
  gghighlight(species_code == "VIRO") + 
  labs(x = "Mean leaf N content (%)", y = NULL) + 
  theme_bw() 
```

## Stem width 

```{r plot mean stem width for each site, fig.height=4, fig.width=4}
stature_tidy %>%
  ggplot(aes(y = species_code, x = mean_stem)) +
  geom_point() + 
  gghighlight(species_code == "VIRO") + 
  facet_wrap(~site_code) + 
  labs(x = "Mean height (cm)", y = NULL) +
  theme_bw() +
  theme(axis.text.y = element_blank())
```
```{r plot mean stem width across all sites, fig.height=4, fig.width=4}
stature_tidy2 %>%
  mutate(species_code = factor(species_code),
         species_code = fct_reorder(species_code, mean_stem)) %>%
  ggplot(aes(y = species_code, x = mean_stem)) +
  geom_point() + 
  gghighlight(species_code == "VIRO") + 
  labs(x = "Mean stem width (nm)", y = NULL) +
  theme_bw() 
```
## Number of Leaves

```{r plot mean number of leaves at each site, fig.height=4, fig.width=4}

stature_tidy %>%
  ggplot(aes(y = species_code, x = mean_num_leaves)) +
  geom_point() + 
  gghighlight(species_code == "VIRO") + 
  facet_wrap(~site_code) + 
  labs(x = "Mean number of leaves", y = NULL) +
  theme_bw() +
  theme(axis.text.y = element_blank())
  
```

```{r plot mean number of leaves across all sites, fig.height=4, fig.width=4, message=FALSE}
stature_tidy2 %>%
  mutate(species_code = factor(species_code),
         species_code = fct_reorder(species_code, mean_num_leaves)) %>%
  ggplot(aes(y = species_code, x = mean_num_leaves)) +
  geom_point() + 
  gghighlight(species_code == "VIRO") + 
  labs(x = "Mean number of leaves", y = NULL) +
  theme_bw() 
```


# Explotatory analysis: PCA

```{r clean trait data into a single df}
traits <- CN_tidy2 %>%
  inner_join(leaf_tidy2, by = "species_code") %>%
  inner_join(stature_tidy2, by = "species_code") %>%
  select(
    species_code, 
    mean_leaf_C,
    mean_leaf_N,
    mean_ldmc, 
    mean_height, 
    mean_stem, 
    mean_num_leaves,
    mean_sla
  ) 

traits2 <- select(traits, -species_code)
```

```{r}
traits_pca <- prcomp(traits2, center = TRUE, scale = TRUE)  

# scree plot
plot(traits_pca)
```

```{r PC1 traits, fig.height=5, fig.width=5}
data.frame(
  trait = rownames(traits_pca$rotation),
  PC1  = traits_pca$rotation[,1],
  PC2  = traits_pca$rotation[,2]
  ) %>%
mutate(trait = factor(trait),
       trait = fct_reorder(trait, PC1)) %>%
ggplot(aes(y = trait, x = PC1)) + 
  geom_col() +
  labs(y = NULL) + 
  theme_bw()
```

```{r PC2 traits, fig.height=5, fig.width=5}
data.frame(
  trait = rownames(traits_pca$rotation),
  PC1  = traits_pca$rotation[,1],
  PC2  = traits_pca$rotation[,2]
  ) %>%
mutate(trait = factor(trait),
       trait = fct_reorder(trait, PC2)) %>%
ggplot(aes(y = trait, x = PC2)) + 
  geom_col() +
  labs(y = NULL) + 
  theme_bw()
```

```{r pca ind, fig.height=5, fig.width=5}
data.frame(
  code = traits$species_code,
  PC1  = traits_pca$x[,1],
  PC2  = traits_pca$x[,2]
  ) %>%
  
ggplot(aes(x = PC1, y = PC2, label = code)) + 
  geom_text() + 
  gghighlight(code %in% c("VIRO", "ALPE")) + 
  geom_hline(yintercept = 0, linetype = "dashed") + 
  geom_vline(xintercept = 0, linetype = "dashed") +
  theme_bw()
```

# Phylogenetic tree

```{r}
tree <- phylomatic(
  taxa = plants$SCIENTIFIC_NAME,
  taxnames = TRUE, 
  get = 'POST',
  method = 'phylomatic',
  storedtree = 'zanne2014'
)
```

